---
title: "Plotting"
output: pdf_document
date: "2023-10-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries-setup, warning = F, message = F}
library(tidyverse)
library(showtext)
library(ggtext)
library(ggridges)

df <- read_csv("data/lemurs.csv")
font_add_google("DM Serif Display", "dm")
showtext_auto()
```

# Exploratory Data Analysis

## Shannon Entropy

```{r echo=F,  fig.cap = "Distributions of Shannon Entropy by Species", fig.pos = "!h"}
ggplot(data = df, aes(
  x = shannon_entropy,
  y = species,
  fill = factor(after_stat(quantile))
)) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = 4,
    quantile_lines = TRUE
  ) +
  scale_fill_manual(values = c("#FFF5E4", "#FFE3E1", "#FFD1D1", "#FF9494")) +
  labs(title = "",
       x = "Shannon Entropy") +
  theme_ridges() +
  theme(
    legend.position = "none",
    axis.title.x = element_text(hjust = 0.5), 
    axis.title.y = element_blank(),
    plot.title = element_text(hjust = 0.5),
    text = element_text(family = "dm")
  )
```

```{r shan-ent, echo = F}
new <- c("spring" = "Spring", "summer" = "Summer")

ggplot(data = df, mapping = aes(x = minutes_after_6am,
                                y = shannon_entropy,
                                color = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se=F) +
  geom_rect(aes(xmin= 0, xmax = 360,
                ymin = 5, ymax = 7.5),
            fill = 'red', alpha = 0.5, color = "red") +
  facet_wrap(~season, labeller = labeller(season = new)) + 
  scale_color_manual(values = c("#3acadf", "#8a64d6")) +
  labs(x = "Minutes after 6 AM",
       y = "Shannon Entropy",
       title = "Shannon Entropy Over Time by Season and Species",
       color = "Species") +
  theme_bw() +
  theme(text = element_text(family = "dm"))
```

  - Shows that L. catta tends to see higher Shannon Entropy than P. coquereli in the summer, but this is reversed in the spring. 

## Faith PD

```{r echo=F, fig.cap = "Distributions of Phylogenetic Diversity by Species", fig.pos = "!h"}
ggplot(data = df, aes(
  x = faith_pd,
  y = species,
  fill = factor(after_stat(quantile))
)) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = 4,
    quantile_lines = TRUE
  ) +
  scale_fill_manual(values = c("#ffbaba", "#ff7b7b", "#ff5252", "#ff0000")) +
  labs(title = "",
       x = "Faith's Phylogenetic Diversity") +
  theme_ridges() +
  theme(
    legend.position = "none",
    axis.title.x = element_text(hjust = 0.5), 
    axis.title.y = element_blank(),
    plot.title = element_text(hjust = 0.5),
    text = element_text(family = "dm")
  )
```

```{r faith-pd, echo = F}
ggplot(data = df, mapping = aes(x = minutes_after_6am,
                                y = faith_pd,
                                color = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se=F) +
  facet_wrap(~season, labeller = labeller(season = new)) + 
  scale_color_manual(values = c("#3acadf", "#8a64d6")) +
  labs(x = "Minutes after 6 AM",
       y = "Phylogenetic Diversity",
       title = "Phylogenetic Diversity Over Time by Season and Species",
       color = "Species") +
  theme_bw() +
  theme(text = element_text(family = "dm"))
```

  - Shows that L. catta consistently has higher phylogenetic diversity than P. coquereli, with season seemingly having an insignificant impact.
  
## Observed Features

```{r echo=F, fig.cap = "Distributions of Observed Features by Species", fig.pos = "!h"}
ggplot(data = df, aes(
  x = observed_features,
  y = species,
  fill = factor(after_stat(quantile))
)) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = 4,
    quantile_lines = TRUE
  ) +
  scale_fill_manual(values = c("#E7FFCE", "#D8FFB1", "#B6DA9F", "#A8D38D")) +
  labs(title = "",
       x = "Number of Observed Features") +
  theme_ridges() +
  theme(
    legend.position = "none",
    axis.title.x = element_text(hjust = 0.5), 
    axis.title.y = element_blank(),
    plot.title = element_text(hjust = 0.5),
    text = element_text(family = "dm")
  )
```

```{r obs-feat, echo = F}
ggplot(data = df, mapping = aes(x = minutes_after_6am,
                                y = observed_features,
                                color = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se=F) +
  facet_wrap(~season, labeller = labeller(season = new)) + 
  scale_color_manual(values = c("#3acadf", "#8a64d6")) +
  labs(x = "Minutes after 6 AM",
       y = "Observed Features",
       title = "Observed Features Over Time by Season and Species",
       color = "Species") +
  theme_bw() +
  theme(text = element_text(family = "dm"))
```

  - Shows that L. catta consistently has more observed features than P. coquereli, but this difference seems to be significantly exacerbated in the summer, suggesting a seasonal impact.

# Initial Exploration

```{r init, echo = F}
#Top microbiomes and their composition in each sample
ggplot(df, aes(x = AMPM, y = shannon_entropy)) +
  geom_boxplot()

ggplot(df, aes(x = AMPM, y = faith_pd)) +
  geom_boxplot()

ggplot(df, aes(x = AMPM, y = observed_features)) +
  geom_boxplot()

hist(df$shannon_entropy)
df.AM <- df %>% 
  filter(AMPM == 'AM')
df.PM <- df %>% 
  filter(AMPM == 'PM')

t.test(df.AM$shannon_entropy, df.PM$shannon_entropy)

#By continuous time
ggplot(df, aes(x = Sampline.time, y = shannon_entropy)) +
  geom_point()

ggplot(df, aes(x = Sampline.time, y = faith_pd)) +
  geom_smooth()

ggplot(df, aes(x = Sampline.time, y = observed_features)) +
  geom_smooth()

df$Sampling.date = as.Date(df$Sampling.date, format = "%d-%b-%Y")


# Shannon Entropy by Species Over time 
df %>%
  select(c(species, 
           Animal.ID, 
           Group, 
           season, 
           Sampling.date, 
           AMPM, 
           Sampline.time, 
           shannon_entropy, 
           observed_features, 
           faith_pd)) %>%
  mutate(Sampling.month = month(Sampling.date)) %>%
  ggplot(aes(x = Sampling.date, y = shannon_entropy, color = species)) + 
  geom_point()


# Shannon Entropy by Social Group Over time 
df %>%
  select(c(species, 
           Animal.ID, 
           Group, 
           season, 
           Sampling.date, 
           AMPM, 
           Sampline.time, 
           shannon_entropy, 
           observed_features, 
           faith_pd)) %>%
  mutate(Sampling.month = month(Sampling.date)) %>%
  ggplot(aes(x = Sampling.date, y = shannon_entropy, color = Group)) + 
  geom_point()

# Faith PD by Species Over time 
df %>%
  select(c(species, 
           Animal.ID, 
           Group, 
           season, 
           Sampling.date, 
           AMPM, 
           Sampline.time, 
           shannon_entropy, 
           observed_features, 
           faith_pd)) %>%
  mutate(Sampling.month = month(Sampling.date)) %>%
  ggplot(aes(x = Sampling.date, y = faith_pd, color = species)) + 
  geom_point()


# Faith PD by Social Group Over time 
df %>%
  select(c(species, 
           Animal.ID, 
           Group, 
           season, 
           Sampling.date, 
           AMPM, 
           Sampline.time, 
           shannon_entropy, 
           observed_features, 
           faith_pd)) %>%
  mutate(Sampling.month = month(Sampling.date)) %>%
  ggplot(aes(x = Sampling.date, y = faith_pd, color = Group)) + 
  geom_point()


# Observed Features by Species Over time 
df %>%
  select(c(species, 
           Animal.ID, 
           Group, 
           season, 
           Sampling.date, 
           AMPM, 
           Sampline.time, 
           shannon_entropy, 
           observed_features, 
           faith_pd)) %>%
  mutate(Sampling.month = month(Sampling.date)) %>%
  ggplot(aes(x = Sampling.date, y = observed_features, color = species)) + 
  geom_point()


# Observed Features by Social Group Over time 
df %>%
  select(c(species, 
           Animal.ID, 
           Group, 
           season, 
           Sampling.date, 
           AMPM, 
           Sampline.time, 
           shannon_entropy, 
           observed_features, 
           faith_pd)) %>%
  mutate(Sampling.month = month(Sampling.date)) %>%
  ggplot(aes(x = Sampling.date, y = observed_features, color = Group)) + 
  geom_point()

# Faith's PD & Shannon Index by Social Group
df %>%
  select(c(species, 
           Animal.ID, 
           Group, 
           season, 
           Sampling.date, 
           AMPM, 
           Sampline.time, 
           shannon_entropy, 
           observed_features, 
           faith_pd)) %>%
  mutate(Sampling.month = month(Sampling.date)) %>%
  ggplot(aes(x = shannon_entropy, y = faith_pd, color = Group)) + 
  geom_point()

# Faith's PD & Shannon Index by Species
df %>%
  select(c(species, 
           Animal.ID, 
           Group, 
           season, 
           Sampling.date, 
           AMPM, 
           Sampline.time, 
           shannon_entropy, 
           observed_features, 
           faith_pd)) %>%
  mutate(Sampling.month = month(Sampling.date)) %>%
  ggplot(aes(x = shannon_entropy, y = faith_pd, color = species)) + 
  geom_point()

# Box plot of observed features in Groups facet by Season fill by AMPM
# Group can also be swapped out for species
df %>%
  select(c(species, 
           Animal.ID, 
           Group, 
           season, 
           Sampling.date, 
           AMPM, 
           Sampline.time, 
           shannon_entropy, 
           observed_features, 
           faith_pd)) %>%
  ggplot(aes(x = Group, y = observed_features, fill = AMPM)) + 
  geom_boxplot() + 
  facet_wrap(~season)

# Box plot of shannon entropy in Groups facet by Season fill by AMPM
# Group can also be swapped out for species
df %>%
  select(c(species, 
           Animal.ID, 
           Group, 
           season, 
           Sampling.date, 
           AMPM, 
           Sampline.time, 
           shannon_entropy, 
           observed_features, 
           faith_pd)) %>%
  ggplot(aes(x = Group, y = shannon_entropy, fill = AMPM)) + 
  geom_boxplot() + 
  facet_wrap(~season)

# Box plot of faith PD in Groups facet by Season fill by AMPM
# Group can also be swapped out for species
df %>%
  select(c(species, 
           Animal.ID, 
           Group, 
           season, 
           Sampling.date, 
           AMPM, 
           Sampline.time, 
           shannon_entropy, 
           observed_features, 
           faith_pd)) %>%
  ggplot(aes(x = Group, y = observed_features, fill = AMPM)) + 
  geom_boxplot() + 
  facet_wrap(~season)


```
```{r}
df |>
  filter(species == "L. catta") |>
  select(starts_with("d__")) |>
  summarise_all(mean, na.rm = TRUE) |>
  pivot_longer(everything(), names_to = "Column", values_to = "Mean_Value") |>
  arrange(Mean_Value) |>
  View()
```

```{r}
family_data <- df |>
  pivot_longer(cols = contains("d__"), names_to = "Bacteria", values_to = "Value") |>
  mutate(Family = str_extract(Bacteria, "(?<=f__)[^;]+")) |>
  mutate(Family = if_else(is.na(Family), "Undefined", Family))
```


```{r message = F}
catta_data <- family_data |>
  filter(species == "L. catta") |>
  group_by(SampleID, Family) |>
  summarize(FamilySum = sum(Value)) |>
  top_n(8) |>
  ungroup() |>
  count(Family) |>
  mutate(PercInTop10Percent = n / nrow(df |> filter(species == "L. catta")),
         Species = "L. catta") |>
  arrange(desc(PercInTop10Percent))

coquereli_data <- family_data |>
  filter(species == "P. coquereli") |>
  group_by(SampleID, Family) |>
  summarize(FamilySum = sum(Value)) |>
  top_n(8) |>
  ungroup() |>
  count(Family) |>
  mutate(PercInTop10Percent = n / nrow(df |> filter(species == "P. coquereli")),
         Species = "P. coquereli") |>
  arrange(desc(PercInTop10Percent))
```

```{r}
plot_order <- catta_data |>
  full_join(coquereli_data, by = join_by(Family == Family)) |>
  select(Family, PercInTop10Percent.x, PercInTop10Percent.y)

plot_order[is.na(plot_order)] <- 0

plot_order <- plot_order |>
  mutate(Difference = abs(PercInTop10Percent.x - PercInTop10Percent.y)) |>
  arrange(desc(Difference)) |>
  slice(1:20)

# family_plot <- plot_order |>
#   pivot_longer(cols = contains("Perc"), names_to = "Species", values_to = "Value") |>
#   mutate(Species = if_else(Species == "PercInTop10Percent.x", "L. catta", "P. coquereli"))

plot_order$Family = factor(plot_order$Family)
plot_order$Family <-
  reorder(plot_order$Family, plot_order$Difference)

ggplot(data = plot_order) +
  geom_point(
    aes(
      x = PercInTop10Percent.x,
      y = Family,
      color = "L. catta"
      ),
    size = 2.5
    ) +
  geom_point(
    aes(
      x = PercInTop10Percent.y,
      y = Family,
      color = "P. coquereli"
      ),
    size = 2.5
    ) +
  geom_segment(
    aes(
      x = PercInTop10Percent.x,
      xend = PercInTop10Percent.y,
      y = Family,
      yend = Family
    ),
    color = "black"
  ) +
  scale_color_manual(values = c("#3acadf", "#8a64d6")) +
  scale_x_continuous(labels = scales::percent) +
  labs(color = "Species", x = "Top 10% Prevalence") +
  theme_bw() +
  theme(text = element_text(family = "dm"))
```

